---
title: "Installing R and RStudio"
---

## Introduction

R is a powerful programming language and free software environment for statistical computing and graphics. In this guide, we'll walk you through installing R and RStudio, which together provide everything you need to start your journey in data analysis and statistical computing.

## Why R and RStudio?

Before diving into installation, let's understand what we're installing:

**R** is the core programming language that provides:

- Comprehensive statistical and graphical capabilities
- Extensive package ecosystem (CRAN)
- Active community and continuous development
- Free and open-source

**RStudio** is an Integrated Development Environment (IDE) that makes working with R much easier by providing:

- Syntax highlighting and code completion
- Integrated help and documentation
- Project management tools
- Built-in plotting and data viewing
- Package management interface

Think of R as the engine and RStudio as the comfortable cockpit that makes driving the engine a pleasant experience.

## Installation Steps

### Step 1: Install R

R must be installed first, as RStudio depends on it.

#### For Windows:

1. Go to the [CRAN website](https://cran.r-project.org/)
2. Click "Download R for Windows"
3. Click "base" (for first-time users)
4. Download the latest version (currently R 4.4.x)
5. Run the downloaded installer and follow the prompts
6. Use default settings unless you have specific preferences

#### For macOS:

1. Go to the [CRAN website](https://cran.r-project.org/)
2. Click "Download R for macOS"
3. Download the appropriate installer:
   - For Apple Silicon Macs (M1/M2): R-4.4.x-arm64.pkg
   - For Intel Macs: R-4.4.x-x86_64.pkg
4. Open the downloaded .pkg file and follow the installation steps

#### For Linux:

Most Linux distributions include R in their package repositories. For Ubuntu/Debian:

```bash
sudo apt update
sudo apt install r-base r-base-dev
```

For more detailed Linux instructions, visit the [CRAN Linux page](https://cran.r-project.org/bin/linux/).

### Step 2: Install RStudio Desktop

After installing R successfully:

1. Go to the [RStudio download page](https://posit.co/download/rstudio-desktop/)
2. Scroll down to "All Installers and Tarballs"
3. Choose the appropriate installer for your operating system:
   - Windows: `.exe` installer
   - macOS: `.dmg` installer
   - Linux: `.deb` or `.rpm` package
4. Download and run the installer
5. Follow the installation prompts

## Verifying Your Installation

Let's make sure everything is working correctly:

### Test R Installation

1. Open RStudio (it should automatically find your R installation)
2. In the Console pane (usually bottom-left), you should see something like:

```bash
R version 4.4.1 (2024-06-14) -- "Race for Your Life"
Copyright (C) 2024 The R Foundation for Statistical Computing
Platform: x86_64-pc-linux-gnu (64-bit)
...
Type 'q()' to quit R.

>
```

3. Try a simple command:

```{r}
#| eval: false
# Test basic arithmetic
2 + 2

# Check R version
R.version.string

# Get help
help(mean)
```

If these work without errors, congratulations! Your R installation is successful.

## Essential Configuration

### Setting Your Working Directory

It's helpful to organize your R work in a dedicated folder:

```{r}
#| eval: false
# Check current working directory
getwd()

# Set working directory (adjust path as needed)
setwd("~/Documents/R-Projects")

# Or better yet, use RStudio Projects (covered in next module)
```

### Installing Essential Packages

R's power comes from its packages. Let's install some essential ones:

```{r}
#| eval: false
# Install the tidyverse (collection of data science packages)
install.packages("tidyverse")

# Install other useful packages
install.packages(c("here", "rmarkdown", "devtools"))

# Load a package to test
library(tidyverse)
```

## Package Management Fundamentals

Understanding package management is crucial for R success:

### CRAN Packages

```{r}
#| eval: false
# Install from CRAN (most common)
install.packages("package_name")

# Install multiple packages
install.packages(c("dplyr", "ggplot2", "readr"))

# Update packages
update.packages()

# Remove packages
remove.packages("package_name")
```

### Loading Packages

```{r}
#| eval: false
# Load package for current session
library(dplyr)

# Alternative loading method
require(dplyr)

# Check installed packages
installed.packages()

# Check loaded packages
search()
```

## RStudio Layout Overview

When you first open RStudio, you'll see four panes:

1. **Console** (bottom-left): Where R commands are executed
2. **Source** (top-left): Text editor for scripts and documents
3. **Environment/History** (top-right): Variables and command history
4. **Files/Plots/Packages/Help** (bottom-right): File browser, plots, package manager, and help

Don't worry if this seems overwhelming – we'll explore each pane in detail in the next lesson!

## Common Installation Issues

### "R not found" Error in RStudio

- **Solution**: Reinstall R first, then RStudio
- RStudio needs to detect your R installation

### Package Installation Fails

- **Windows**: Install Rtools from [CRAN](https://cran.r-project.org/bin/windows/Rtools/)
- **macOS**: Install Xcode command line tools: `xcode-select --install`
- **All platforms**: Try installing packages from a different mirror

### Permission Issues

- **Solution**: Run RStudio as administrator (Windows) or use `sudo` (Linux/macOS) for package installation

## Alternative Installation Methods

### Using Package Managers

#### Windows - Chocolatey:
```bash
choco install r.project rstudio
```

#### macOS - Homebrew:
```bash
brew install --cask r rstudio
```

#### Linux - Conda:
```bash
conda install -c r r-base rstudio
```

## Next Steps

Now that you have R and RStudio installed:

1. Explore the [RStudio Interface](rstudio-interface.qmd)
2. Try [R as a Calculator](r-as-calculator.qmd)
3. Learn about [R Scripts and Projects](r-scripts-projects.qmd)

## Resources for Further Learning

- [RStudio Education](https://education.rstudio.com/)
- [R for Data Science book](https://r4ds.had.co.nz/) (free online)
- [CRAN Task Views](https://cran.r-project.org/web/views/)
- [RStudio Cheat Sheets](https://rstudio.com/resources/cheatsheets/)

::: {.callout-tip}
## Pro Tip

Create a dedicated folder for your R projects and always use RStudio Projects to manage your work. This keeps your files organized and makes your analysis reproducible!
:::

## Congratulations!

You've successfully set up your R environment! You're now ready to start your data science journey with R.
