---
title: "Week 7"
format: html
subtitle: "Communities, Roles, and Positions in Networks"
date: "today"
callout-appearance: "default"
---

## Overview

This week shifts focus from individual nodes and local structures to
understanding networks at a higher level of abstraction. We explore two
complementary approaches to seeing social structure:

1. **Connectionist approaches**: How cohesive communities form through direct
   ties and resource flows
2. **Positional approaches**: How roles and positions emerge through similar
   patterns of ties

These perspectives help us address fundamental questions about social
integration: What holds groups together? How do we identify meaningful
subgroups? How do role structures organize social life?

## Prepare

::: {.callout-note}
## 📚

📖 Review the [syllabus](../../course/syllabus.qmd)

📖 Review concepts from previous weeks:

- [Network fundamentals](../week-1/network_terms.qmd) (Week 1)
- [Centrality measures](../week-3/slideshow.html) (Week 3) 
- [Triadic structures](../week-4/slideshow.html) (Week 4)

📖 **Required readings** - Book chapters available in this directory:

Chapter 8: "Cohesion and Groups" and  Chapter 10: "Positions and Roles" from
Rawlings, Craig M., Jeffrey A. Smith, James Moody, and Daniel A. McFarland.
[Network analysis: integrating social network theory, method, and application
with R](https://go.exlibris.link/5r30tSdj). Cambridge University Press, 2023.

:::

## Participate

::: {.callout-note}
## 👥

📉 Slideshow: [Communities, Roles, and Positions](slideshow.html)

**Key topics covered**:

**Theoretical Foundations**:
- Durkheim's mechanical vs. organic solidarity
- Granovetter's strength of weak ties
- Watts-Strogatz small-world networks
- Homophily and social integration

**Part I: Network Communities**

- Social cohesion and integration
- Measuring connectivity: density, clustering, path lengths
- Small-world networks and the strength of weak ties
- **Understanding modularity**: formula, interpretation, thresholds
- Community detection algorithms:
  - Louvain and Leiden (modularity-based)
  - Walktrap (random walk-based)
  - Edge betweenness (Girvan-Newman)
  - Label propagation
- **Algorithm comparison**: when to use each method
- **Resolution parameter**: controlling community granularity
- Assessing community structure and fit

**Part II: Roles and Positions**

- From actors to roles: the dramaturgical metaphor
- **Types of structural equivalence**: structural, regular, automorphic
- **Measuring equivalence**: correlation and distance methods
- Blockmodeling procedures and image matrices
- **CONCOR algorithm**: convergence of iterated correlations
- Hierarchical cluster analysis
- Multi-relational role structures
- Ideal role types and assessing fit

**Integration: Communities ≠ Positions**

- Comparing connectionist and positional approaches
- When partitions align vs. diverge
- Adjusted Rand Index for measuring overlap
- Methodological pluralism

:::


## Practice


::: {.callout-note}
## 💻

### Tutorial 1: [Community Detection in R](practice_1.qmd)

A comprehensive tutorial covering:

- **Multiple algorithms**: Louvain, Edge Betweenness, Walktrap, Spectral, Fast Greedy
- **Method comparison**: Calculate and compare modularity scores across methods
- **Agreement analysis**: Use Normalized Mutual Information (NMI) to assess method consistency
- **Quality assessment**: Analyze within vs. between community density
- **Resolution exploration**: Sweep resolution parameters to find optimal granularity
- **Stability testing**: Assess robustness to network perturbations
- **Visualization**: Create publication-quality community structure plots
- **Hands-on examples**: Karate Club network and scale-free networks

### Tutorial 2: [Community Detection in Python](practice_2.qmd)

A parallel tutorial using Python and NetworkX that demonstrates:

- Loading and exploring networks in Python
- Applying greedy modularity, label propagation, and Girvan-Newman algorithms
- Optionally running Louvain (if `python-louvain` is installed)
- Computing modularity using `networkx.algorithms.community`
- Visualizing communities with Matplotlib
- Comparing methods via a modularity summary table and bar plot
- Applying methods to both the karate club graph and a larger scale-free network

### Tutorial 3: [Role and Position Analysis in R](practice_3.qmd)

A comprehensive tutorial covering blockmodeling and structural equivalence:

- **Structural equivalence measures**: Correlation and Euclidean distance
- **CONCOR algorithm**: Classical blockmodeling approach
- **Hierarchical clustering**: Identifying positions from tie profiles
- **Blockmodels and image matrices**: Reduced network representations
- **Block density analysis**: Within vs. between position densities
- **Multiple methods comparison**: Comparing different equivalence measures
- **Model fit assessment**: Evaluating blockmodel quality
- **Practical examples**: Corporate advice networks and organizational hierarchies
- **Substantive interpretation**: Understanding role meanings

### Advanced Topics (Coming Soon)

- Small-world analysis and metrics
- Cohesive blocking and k-components
- Dynamic community detection
- Overlapping community detection

:::


## Perform


::: {.callout-note}
## 📝

**Application exercises**:

1. **Community Analysis**: Apply multiple community detection algorithms to a
   network and compare results. Assess which method best captures substantive
communities.

2. **Role Structure Analysis**: Perform a blockmodeling analysis to identify
   role positions in an organizational or social network. Create and interpret
the image matrix.

3. **Integration**: Examine whether communities correspond to positions - do
   structurally equivalent actors tend to be in the same community?

:::

## Ponder

::: {.callout-note}
## 🤔

### Core Readings

**Cohesion and Communities**

Chapter 8 ("Cohesion and Groups") from Borgatti, S. P., Everett, M. G.,
Johnson, J. C., & Agneessens, F. (2024). *Analyzing Social Networks Using R*.
Sage Publications.

**Foundational Work on Blockmodeling**

White, H. C., Boorman, S. A., & Breiger, R. L. (1976). Social structure from
multiple networks. I. Blockmodels of roles and positions. *American Journal of
Sociology*, 81(4), 730-780.

**Community Detection Algorithms**

Newman, M. E. J., & Girvan, M. (2004). Finding and evaluating community
structure in networks. *Physical Review E*, 69(2), 026113.

Blondel, V. D., Guillaume, J. L., Lambiotte, R., & Lefebvre, E. (2008). Fast
unfolding of communities in large networks. *Journal of Statistical Mechanics:
Theory and Experiment*, 2008(10), P10008.

**Small-World Networks**

Watts, D. J., & Strogatz, S. H. (1998). Collective dynamics of 'small-world'
networks. *Nature*, 393(6684), 440-442.

Granovetter, M. S. (1973). The strength of weak ties. *American Journal of
Sociology*, 78(6), 1360-1380.

### Supplementary Readings

**Structural Cohesion**

Moody, J., & White, D. R. (2003). Structural cohesion and embeddedness: A
hierarchical concept of social groups. *American Sociological Review*, 68(1),
103-127.

**Generalized Blockmodeling**

Doreian, P., Batagelj, V., & Ferligoj, A. (2005). *Generalized Blockmodeling*.
Cambridge University Press.

**Role Equivalence**

Burt, R. S. (1990). Detecting role equivalence. *Social Networks*, 12(1),
83-97.

**Applications**

Padgett, J. F., & Ansell, C. K. (1993). Robust action and the rise of the
Medici, 1400-1434. *American Journal of Sociology*, 98(6), 1259-1319.

### Conceptual Connections

**Mechanical vs. Organic Solidarity**

Consider how Durkheim's distinction between mechanical solidarity (based on
similarity) and organic solidarity (based on differentiated interdependence)
maps onto the distinction between communities and role structures.

**Methodological Pluralism**

Reflect on why there is no single "best" method for community detection or
blockmodeling. What does this tell us about the nature of social structure?

**Integration of Approaches**

How might communities and positions relate to each other? Can structurally
equivalent actors be in different communities? Can community members occupy
different positions?

:::


## Quick Reference: Essential R Code

### Community Detection Snippets

```{r}
#| label: reference-communities
#| eval: false
#| code-fold: show

library(igraph)

# Load network
g <- make_graph("Zachary")  # or read from file

# Method 1: Louvain (fast, modularity-based)
comm_louvain <- cluster_louvain(g)

# Method 2: Leiden (improved Louvain)
comm_leiden <- cluster_leiden(g)

# Method 3: Walktrap (random walk-based)
comm_walktrap <- cluster_walktrap(g, steps = 4)

# Method 4: Edge Betweenness (hierarchical)
comm_edgebet <- cluster_edge_betweenness(g)

# Method 5: Fast Greedy (hierarchical modularity)
comm_fastgreedy <- cluster_fast_greedy(g)

# Extract results
membership(comm_louvain)         # Community assignments
length(comm_louvain)             # Number of communities
sizes(comm_louvain)              # Community sizes
modularity(comm_louvain)         # Modularity score

# Compare methods
compare(comm_louvain, comm_leiden, method = "nmi")  # Agreement

# Visualize
plot(comm_louvain, g,
     vertex.size = 8,
     edge.color = "gray80",
     main = "Communities")
```

### Blockmodeling Snippets

```{r}
#| label: reference-blockmodels
#| eval: false
#| code-fold: show

library(igraph)

g <- sample_pa(20, directed = TRUE)
adj <- as.matrix(as_adjacency_matrix(g))

# Method 1: Correlation-based equivalence
equiv_cor <- cor(t(adj))          # Correlate row profiles
dist_cor <- as.dist(1 - equiv_cor)
hc_cor <- hclust(dist_cor, method = "complete")
positions_cor <- cutree(hc_cor, k = 4)

# Method 2: Euclidean distance equivalence
dist_euclidean <- dist(adj, method = "euclidean")
hc_euclidean <- hclust(dist_euclidean, method = "ward.D2")
positions_euclidean <- cutree(hc_euclidean, k = 4)

# Visualize dendrogram
plot(hc_cor, main = "Position Hierarchy", xlab = "Node")
rect.hclust(hc_cor, k = 4, border = "red")

# Create image matrix
n_pos <- max(positions_cor)
image_matrix <- matrix(0, nrow = n_pos, ncol = n_pos)

for(i in 1:n_pos) {
  for(j in 1:n_pos) {
    block <- adj[positions_cor == i, positions_cor == j]
    image_matrix[i, j] <- mean(block)
  }
}

print(round(image_matrix, 2))

# Visualize with position colors
V(g)$color <- rainbow(n_pos)[positions_cor]
plot(g, vertex.size = 8, main = "Network by Position")
```

### Essential Network Data Sources

```{r}
#| label: reference-data
#| eval: false
#| code-fold: show

library(igraph)

# Built-in networks
karate <- make_graph("Zachary")              # Karate club
dolphins <- make_graph("Dolphins")            # Dolphin social network

# Generate synthetic networks
erdos_renyi <- sample_gnp(100, 0.05)         # Random
barabasi <- sample_pa(100, power = 1)        # Scale-free
small_world <- watts.strogatz.game(1, 100, 5, 0.1)  # Small-world

# Read from file
g_edgelist <- read_graph("network.txt", format = "edgelist")
g_graphml <- read_graph("network.graphml", format = "graphml")
g_gml <- read_graph("network.gml", format = "gml")

# From adjacency matrix
adj_matrix <- as.matrix(read.csv("adjacency.csv", row.names = 1))
g_adj <- graph_from_adjacency_matrix(adj_matrix, mode = "undirected")

# From data frame
edges_df <- data.frame(from = c("A", "B", "C"),
                       to = c("B", "C", "D"))
g_df <- graph_from_data_frame(edges_df, directed = TRUE)
```

## Additional Resources


::: {.callout-tip}
## 🔧 Software and Tools

**R packages**:

- `igraph`: Community detection, blockmodeling
- `sna` and `network`: Role equivalence, structural cohesion
- `blockmodeling`: Generalized blockmodeling

**Python packages**:

- `igraph`: Community detection algorithms
- `networkx`: Modularity, community detection
- `python-louvain`: Louvain algorithm implementation
- `leidenalg`: Leiden algorithm

**Visualization**:

- Gephi: Interactive community visualization
- Cytoscape: Network analysis and visualization
- R packages: `ggraph`, `visNetwork`

:::
