---
title: "Community Detection in R"
subtitle: "Identifying cohesive subgroups in networks"
format: 
  html:
    code-fold: false
    toc: true
    toc-depth: 3
    number-sections: true
execute:
  warning: false
  message: false
---

# Introduction

This tutorial demonstrates how to detect and analyze communities in networks using R. We'll explore multiple algorithms, compare their results, and learn how to assess community structure quality.

## Learning Objectives

By the end of this tutorial, you will be able to:

- Implement multiple community detection algorithms in R
- Calculate and interpret modularity scores
- Compare community detection methods
- Visualize community structure
- Assess the quality and robustness of detected communities

## Required Packages

```{r}
#| label: setup

# Load required packages
library(igraph)
library(ggplot2)
library(dplyr)
library(tidyr)
library(patchwork)

# Set seed for reproducibility
set.seed(42)
```

# Loading and Exploring Network Data

## Example 1: Karate Club Network

We'll start with the classic Zachary's Karate Club network - a well-studied network documenting social interactions in a karate club that eventually split into two groups.

```{r}
#| label: load-karate

# Load the karate club network
karate <- make_graph("Zachary")

# Basic network statistics
cat("Network size:", vcount(karate), "nodes\n")
cat("Network edges:", ecount(karate), "edges\n")
cat("Network density:", edge_density(karate), "\n")
cat("Average clustering:", transitivity(karate, type = "average"), "\n")
```

## Visualizing the Original Network

```{r}
#| label: viz-original
#| fig-width: 8
#| fig-height: 6

# Set layout for consistent visualization
set.seed(42)
layout_karate <- layout_with_fr(karate)

# Plot the network
plot(karate,
     layout = layout_karate,
     vertex.size = 10,
     vertex.label.cex = 0.7,
     vertex.color = "lightblue",
     vertex.frame.color = "darkblue",
     edge.color = "gray80",
     main = "Zachary's Karate Club Network")
```

# Community Detection Algorithms

## Method 1: Louvain Algorithm

The Louvain algorithm is a fast, hierarchical method that maximizes modularity.

```{r}
#| label: louvain

# Detect communities using Louvain
louvain_comm <- cluster_louvain(karate)

# Summary statistics
cat("Number of communities:", length(louvain_comm), "\n")
cat("Modularity:", modularity(louvain_comm), "\n")
cat("Community sizes:", sizes(louvain_comm), "\n")

# Community membership
print(membership(louvain_comm))
```

### Visualizing Louvain Communities

```{r}
#| label: viz-louvain
#| fig-width: 8
#| fig-height: 6

# Define colors for communities
colors <- c("coral", "lightblue", "lightgreen", "gold", "plum")

plot(louvain_comm, karate,
     layout = layout_karate,
     vertex.size = 10,
     vertex.label.cex = 0.7,
     edge.color = "gray80",
     main = paste0("Louvain Algorithm\nModularity = ", 
                   round(modularity(louvain_comm), 3)))
```

## Method 2: Edge Betweenness

The edge betweenness algorithm identifies communities by removing edges with high betweenness.

```{r}
#| label: edge-betweenness

# Detect communities using edge betweenness
eb_comm <- cluster_edge_betweenness(karate)

cat("Number of communities:", length(eb_comm), "\n")
cat("Modularity:", modularity(eb_comm), "\n")
cat("Community sizes:", sizes(eb_comm), "\n")
```

### Visualizing Edge Betweenness Communities

```{r}
#| label: viz-edge-betweenness
#| fig-width: 8
#| fig-height: 6

plot(eb_comm, karate,
     layout = layout_karate,
     vertex.size = 10,
     vertex.label.cex = 0.7,
     edge.color = "gray80",
     main = paste0("Edge Betweenness Algorithm\nModularity = ", 
                   round(modularity(eb_comm), 3)))
```

## Method 3: Walktrap

The walktrap algorithm uses random walks to identify communities.

```{r}
#| label: walktrap

# Detect communities using walktrap
walktrap_comm <- cluster_walktrap(karate)

cat("Number of communities:", length(walktrap_comm), "\n")
cat("Modularity:", modularity(walktrap_comm), "\n")
cat("Community sizes:", sizes(walktrap_comm), "\n")
```

### Visualizing Walktrap Communities

```{r}
#| label: viz-walktrap
#| fig-width: 8
#| fig-height: 6

plot(walktrap_comm, karate,
     layout = layout_karate,
     vertex.size = 10,
     vertex.label.cex = 0.7,
     edge.color = "gray80",
     main = paste0("Walktrap Algorithm\nModularity = ", 
                   round(modularity(walktrap_comm), 3)))
```

## Method 4: Leading Eigenvector (Spectral)

This method uses the leading eigenvector of the modularity matrix.

```{r}
#| label: spectral

# Detect communities using leading eigenvector
spectral_comm <- cluster_leading_eigen(karate)

cat("Number of communities:", length(spectral_comm), "\n")
cat("Modularity:", modularity(spectral_comm), "\n")
cat("Community sizes:", sizes(spectral_comm), "\n")
```

### Visualizing Spectral Communities

```{r}
#| label: viz-spectral
#| fig-width: 8
#| fig-height: 6

plot(spectral_comm, karate,
     layout = layout_karate,
     vertex.size = 10,
     vertex.label.cex = 0.7,
     edge.color = "gray80",
     main = paste0("Spectral Method\nModularity = ", 
                   round(modularity(spectral_comm), 3)))
```

## Method 5: Fast Greedy

A hierarchical agglomerative algorithm that's fast for large networks.

```{r}
#| label: fast-greedy

# Detect communities using fast greedy
fg_comm <- cluster_fast_greedy(karate)

cat("Number of communities:", length(fg_comm), "\n")
cat("Modularity:", modularity(fg_comm), "\n")
cat("Community sizes:", sizes(fg_comm), "\n")
```

# Comparing Methods

## Summary Comparison Table

```{r}
#| label: comparison-table

# Create comparison data frame
comparison <- data.frame(
  Method = c("Louvain", "Edge Betweenness", "Walktrap", 
             "Spectral", "Fast Greedy"),
  Communities = c(length(louvain_comm), length(eb_comm), 
                  length(walktrap_comm), length(spectral_comm), 
                  length(fg_comm)),
  Modularity = c(modularity(louvain_comm), modularity(eb_comm),
                 modularity(walktrap_comm), modularity(spectral_comm),
                 modularity(fg_comm))
)

# Display table
knitr::kable(comparison, 
             digits = 4,
             caption = "Comparison of Community Detection Methods")
```

## Modularity Comparison Plot

```{r}
#| label: modularity-plot
#| fig-width: 8
#| fig-height: 5

ggplot(comparison, aes(x = reorder(Method, Modularity), y = Modularity)) +
  geom_col(fill = "steelblue", alpha = 0.8) +
  geom_text(aes(label = round(Modularity, 3)), 
            vjust = -0.5, size = 4) +
  coord_flip() +
  labs(title = "Modularity Scores by Community Detection Method",
       x = "Method",
       y = "Modularity") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 14))
```

## Agreement Between Methods

Calculate how similar the community assignments are across methods.

```{r}
#| label: method-agreement

# Function to calculate normalized mutual information
calculate_nmi <- function(comm1, comm2) {
  compare(comm1, comm2, method = "nmi")
}

# Create agreement matrix
methods_list <- list(
  Louvain = louvain_comm,
  EdgeBet = eb_comm,
  Walktrap = walktrap_comm,
  Spectral = spectral_comm,
  FastGreedy = fg_comm
)

# Calculate pairwise NMI
nmi_matrix <- matrix(0, nrow = 5, ncol = 5)
rownames(nmi_matrix) <- colnames(nmi_matrix) <- names(methods_list)

for(i in 1:5) {
  for(j in 1:5) {
    nmi_matrix[i,j] <- calculate_nmi(methods_list[[i]], methods_list[[j]])
  }
}

# Display as heatmap
library(reshape2)
nmi_long <- melt(nmi_matrix)

ggplot(nmi_long, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  geom_text(aes(label = round(value, 2)), color = "white", size = 4) +
  scale_fill_gradient2(low = "blue", mid = "purple", high = "red",
                       midpoint = 0.5, limit = c(0, 1)) +
  labs(title = "Agreement Between Methods (NMI)",
       x = "", y = "", fill = "NMI") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(face = "bold", size = 14))
```

# Assessing Community Quality

## Within vs. Between Community Density

```{r}
#| label: density-analysis

# Function to calculate within and between community density
analyze_density <- function(graph, communities) {
  membership <- membership(communities)
  n_comm <- max(membership)
  
  within_density <- numeric(n_comm)
  
  for(i in 1:n_comm) {
    subgraph <- induced_subgraph(graph, which(membership == i))
    within_density[i] <- edge_density(subgraph)
  }
  
  # Calculate between community density
  total_possible <- vcount(graph) * (vcount(graph) - 1) / 2
  within_nodes <- sapply(1:n_comm, function(i) sum(membership == i))
  within_possible <- sum(within_nodes * (within_nodes - 1) / 2)
  between_possible <- total_possible - within_possible
  
  between_edges <- ecount(graph) - 
    sum(sapply(1:n_comm, function(i) {
      ecount(induced_subgraph(graph, which(membership == i)))
    }))
  
  between_density <- between_edges / between_possible
  
  list(
    within_mean = mean(within_density),
    within_sd = sd(within_density),
    between = between_density,
    ratio = mean(within_density) / between_density
  )
}

# Analyze Louvain communities
density_stats <- analyze_density(karate, louvain_comm)

cat("Within-community density (mean):", round(density_stats$within_mean, 3), "\n")
cat("Between-community density:", round(density_stats$between, 3), "\n")
cat("Ratio (within/between):", round(density_stats$ratio, 2), "\n")
```

## Community Size Distribution

```{r}
#| label: size-distribution
#| fig-width: 8
#| fig-height: 5

# Create data frame of community sizes for all methods
size_data <- data.frame(
  Method = rep(names(methods_list), sapply(methods_list, length)),
  Size = unlist(lapply(methods_list, sizes))
)

ggplot(size_data, aes(x = Method, y = Size, fill = Method)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.5) +
  labs(title = "Community Size Distribution by Method",
       x = "Method",
       y = "Community Size") +
  theme_minimal() +
  theme(legend.position = "none",
        plot.title = element_text(face = "bold", size = 14))
```

# Resolution Parameter Exploration

For algorithms that support resolution parameters (like Louvain), we can explore different levels of granularity.

```{r}
#| label: resolution-sweep
#| fig-width: 10
#| fig-height: 6

# Sweep resolution parameter
resolutions <- seq(0.5, 2.0, by = 0.1)
results <- data.frame(
  resolution = numeric(),
  n_communities = numeric(),
  modularity = numeric()
)

for(res in resolutions) {
  comm <- cluster_louvain(karate, resolution = res)
  results <- rbind(results, data.frame(
    resolution = res,
    n_communities = length(comm),
    modularity = modularity(comm)
  ))
}

# Create plots
p1 <- ggplot(results, aes(x = resolution, y = n_communities)) +
  geom_line(color = "steelblue", size = 1.2) +
  geom_point(color = "steelblue", size = 3) +
  labs(title = "Resolution vs Number of Communities",
       x = "Resolution Parameter",
       y = "Number of Communities") +
  theme_minimal()

p2 <- ggplot(results, aes(x = resolution, y = modularity)) +
  geom_line(color = "coral", size = 1.2) +
  geom_point(color = "coral", size = 3) +
  labs(title = "Resolution vs Modularity",
       x = "Resolution Parameter",
       y = "Modularity") +
  theme_minimal()

p1 + p2
```

# Example 2: Larger Network (Random Scale-Free Network)

Let's apply what we've learned to a larger network. We'll generate a scale-free network using the Barabási-Albert model, which is common in many real-world networks.

```{r}
#| label: larger-network
#| fig-width: 10
#| fig-height: 8

# Generate a larger scale-free network
set.seed(42)
large_net <- sample_pa(n = 100, power = 1, m = 2, directed = FALSE)

cat("Network size:", vcount(large_net), "nodes\n")
cat("Network edges:", ecount(large_net), "edges\n")
cat("Network density:", round(edge_density(large_net), 4), "\n")

# Apply Louvain
large_comm <- cluster_louvain(large_net)

cat("\nCommunities detected:", length(large_comm), "\n")
cat("Modularity:", round(modularity(large_comm), 3), "\n")
cat("Community sizes:", sizes(large_comm), "\n")

# Visualize
set.seed(42)
plot(large_comm, large_net,
     vertex.size = 6,
     vertex.label = NA,
     edge.color = "gray90",
     main = paste0("Scale-Free Network Communities (n=100)\n",
                   "Louvain Algorithm, Modularity = ", 
                   round(modularity(large_comm), 3)))
```

# Hierarchical Community Structure

Some algorithms (like edge betweenness and fast greedy) create hierarchical dendrograms.

```{r}
#| label: dendrogram
#| fig-width: 10
#| fig-height: 6

# Use fast greedy to get hierarchy
fg_karate <- cluster_fast_greedy(karate)

# Plot dendrogram
plot(as.dendrogram(fg_karate),
     main = "Hierarchical Community Structure\nFast Greedy Algorithm",
     xlab = "Node",
     ylab = "Merge Height")
```

# Stability Analysis

Test how stable communities are to network perturbations.

```{r}
#| label: stability

# Function to assess stability
assess_stability <- function(graph, n_iterations = 100, 
                             removal_prop = 0.1) {
  
  original_comm <- cluster_louvain(graph)
  original_membership <- membership(original_comm)
  
  similarities <- numeric(n_iterations)
  
  for(i in 1:n_iterations) {
    # Remove random edges
    edges_to_remove <- sample(ecount(graph), 
                               size = floor(ecount(graph) * removal_prop))
    perturbed <- delete_edges(graph, edges_to_remove)
    
    # Detect communities in perturbed network
    perturbed_comm <- cluster_louvain(perturbed)
    
    # Calculate similarity
    similarities[i] <- compare(original_comm, perturbed_comm, 
                               method = "nmi")
  }
  
  similarities
}

# Run stability analysis
stability_scores <- assess_stability(karate, n_iterations = 50)

cat("Mean stability (NMI):", round(mean(stability_scores), 3), "\n")
cat("SD stability:", round(sd(stability_scores), 3), "\n")

# Plot
data.frame(NMI = stability_scores) %>%
  ggplot(aes(x = NMI)) +
  geom_histogram(bins = 20, fill = "steelblue", alpha = 0.7) +
  geom_vline(xintercept = mean(stability_scores), 
             color = "red", linetype = "dashed", size = 1) +
  labs(title = "Community Detection Stability",
       subtitle = "After removing 10% of edges (50 iterations)",
       x = "Normalized Mutual Information",
       y = "Frequency") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 14))
```

# Key Takeaways

1. **Multiple methods exist**: Different algorithms can produce different results. Always compare multiple approaches.

2. **Modularity is key**: Higher modularity generally indicates better community structure, but it's not the only metric to consider.

3. **Context matters**: The "best" method depends on your network properties and research questions.

4. **Resolution matters**: Many algorithms have parameters that control the granularity of detected communities.

5. **Stability is important**: Test how robust your communities are to perturbations.

6. **Visual inspection helps**: Always visualize your communities to ensure they make substantive sense.

# Exercises

## Exercise 1: Compare Methods on Your Own Data

Load a network dataset of your choice and:

1. Apply at least three different community detection methods
2. Compare their modularity scores
3. Visualize the results
4. Assess agreement between methods using NMI

## Exercise 2: Resolution Exploration

Using the Louvain algorithm:

1. Sweep the resolution parameter from 0.3 to 2.5
2. Create a plot showing how the number of communities changes
3. Identify the "elbow point" where adding resolution stops being useful

## Exercise 3: Community Characteristics

For the best community structure you find:

1. Calculate within vs. between community density
2. Identify which nodes are "bridges" between communities
3. Examine node-level attributes (if available) to see if communities correspond to meaningful groups

# Additional Resources

- **igraph documentation**: [Community detection reference](https://igraph.org/r/doc/communities.html)
- **Modularity paper**: Newman & Girvan (2004), *Physical Review E*
- **Louvain algorithm**: Blondel et al. (2008), *Journal of Statistical Mechanics*
- **Comparison of methods**: Fortunato (2010), *Physics Reports*

---

**Next Steps**: After mastering community detection, proceed to `practice_2.qmd` to learn about blockmodeling and role analysis.
