---
title: "Practice Exercise: QAP Regression"
subtitle: "Week 4 - Network Analytics"
format:
  html:
    toc: true
    code-fold: false
---

## Exercise Overview

In this exercise, we'll explore **QAP (Quadratic Assignment Procedure) Regression**, a method for testing relationships between network matrices while accounting for the dependency structure in network data. We'll examine:

1. **QAP Correlation**: Testing association between two networks
2. **QAP Regression**: Predicting one network from multiple others
3. **Interpretation**: Understanding results and significance tests

## Setup

Load required packages:

```{r}
#| message: false
#| warning: false
library(sna)
library(network)
set.seed(456)  # for reproducibility
```

## Part 1: Creating Example Networks

Let's create a scenario with employees in a small company. We'll examine relationships between:
- **Advice network**: Who seeks work advice from whom
- **Friendship network**: Who considers whom a friend  
- **Office proximity**: Physical distance between desks
- **Department**: Same department membership

```{r}
# Create 10 employees
n <- 10
employees <- paste0("Emp", 1:10)

# 1. Create advice network (who seeks advice from whom)
advice <- matrix(0, n, n)
rownames(advice) <- colnames(advice) <- employees

# Senior employees (1-3) give advice to juniors
advice[4:10, 1:3] <- rbinom(21, 1, 0.6)
# Some peer advice among juniors
advice[4:7, 8:10] <- rbinom(12, 1, 0.3)
advice[8:10, 4:7] <- rbinom(12, 1, 0.3)
diag(advice) <- 0  # no self-loops

# 2. Create friendship network (undirected)
friendship <- matrix(0, n, n)
rownames(friendship) <- colnames(friendship) <- employees

# Friends within similar levels
friendship[1:3, 1:3] <- rbinom(9, 1, 0.5)
friendship[4:7, 4:7] <- rbinom(16, 1, 0.4)
friendship[8:10, 8:10] <- rbinom(9, 1, 0.5)
# Some cross-level friendships
friendship[1:3, 4:7] <- rbinom(12, 1, 0.2)
friendship[4:7, 1:3] <- t(friendship[1:3, 4:7])

# Make symmetric and remove diagonal
friendship[lower.tri(friendship)] <- t(friendship)[lower.tri(friendship)]
diag(friendship) <- 0

# 3. Create office proximity (based on desk distance)
# Closer desks = higher values
office_dist <- matrix(0, n, n)
rownames(office_dist) <- colnames(office_dist) <- employees

# Create distance based on floor/section
for(i in 1:n) {
  for(j in 1:n) {
    if(i != j) {
      # Same section (1-3, 4-7, 8-10)
      if((i <= 3 & j <= 3) | 
         (i >= 4 & i <= 7 & j >= 4 & j <= 7) |
         (i >= 8 & j >= 8)) {
        office_dist[i,j] <- runif(1, 0.7, 1)  # close proximity
      } else {
        office_dist[i,j] <- runif(1, 0, 0.3)  # far proximity
      }
    }
  }
}

# 4. Create department membership (same = 1, different = 0)
dept <- c(rep("Sales", 3), rep("Tech", 4), rep("Admin", 3))
same_dept <- matrix(0, n, n)
for(i in 1:n) {
  for(j in 1:n) {
    if(dept[i] == dept[j] & i != j) same_dept[i,j] <- 1
  }
}
rownames(same_dept) <- colnames(same_dept) <- employees
```

## Part 2: Visualizing Networks

Let's visualize our networks to understand their structure:

```{r}
#| fig-width: 12
#| fig-height: 8
par(mfrow = c(2, 2))

# Plot advice network
gplot(advice, 
      vertex.col = c(rep("#c41c85", 3), rep("#50C878", 4), rep("#4169E1", 3)),
      vertex.cex = 2,
      label = employees,
      label.col = "white",
      label.cex = 0.7,
      edge.col = "#D3D3D3",
      main = "Advice Network",
      sub = "(Pink: Senior, Green: Tech, Blue: Admin)")

# Plot friendship network  
gplot(friendship,
      vertex.col = c(rep("#c41c85", 3), rep("#50C878", 4), rep("#4169E1", 3)),
      vertex.cex = 2,
      label = employees,
      label.col = "white", 
      label.cex = 0.7,
      edge.col = "#D3D3D3",
      main = "Friendship Network",
      mode = "fruchtermanreingold")

# Plot office proximity
gplot(office_dist,
      vertex.col = c(rep("#c41c85", 3), rep("#50C878", 4), rep("#4169E1", 3)),
      vertex.cex = 2,
      label = employees,
      label.col = "white",
      label.cex = 0.7,
      edge.col = gray(1 - office_dist),
      edge.lwd = office_dist * 3,
      main = "Office Proximity",
      sub = "(Darker/thicker = closer desks)")

# Plot department membership
gplot(same_dept,
      vertex.col = c(rep("#c41c85", 3), rep("#50C878", 4), rep("#4169E1", 3)),
      vertex.cex = 2,
      label = employees,
      label.col = "white",
      label.cex = 0.7,
      edge.col = "#D3D3D3",
      main = "Same Department")
```

## Part 3: QAP Correlation

Test pairwise correlations between networks using QAP:

```{r}
# Test correlation between advice and friendship
qap_advice_friend <- qaptest(list(advice, friendship), 
                              gcor,
                              g1 = 1, g2 = 2,
                              reps = 1000)

cat("QAP Correlation: Advice ~ Friendship\n")
cat("Observed correlation:", round(qap_advice_friend$testval, 3), "\n")
cat("P-value:", round(qap_advice_friend$pgreq, 3), "\n\n")

# Visualize QAP test
hist(qap_advice_friend$dist, 
     breaks = 30,
     main = "QAP Test: Advice ~ Friendship",
     xlab = "Correlation",
     col = "lightgray")
abline(v = qap_advice_friend$testval, col = "red", lwd = 2)
text(qap_advice_friend$testval, max(table(cut(qap_advice_friend$dist, 30))), 
     "Observed", col = "red", pos = 4)
```

::: {.callout-tip}
## Interpretation
The QAP test permutes one matrix while holding the other fixed, maintaining the dependency structure. If p < 0.05, the correlation is significant.
:::

```{r}
# Test all pairwise correlations
networks <- list(advice, friendship, office_dist, same_dept)
net_names <- c("Advice", "Friendship", "Office", "SameDept")

# Create correlation matrix
qap_cors <- matrix(NA, 4, 4)
qap_pvals <- matrix(NA, 4, 4)

for(i in 1:4) {
  for(j in 1:4) {
    if(i != j) {
      qap_result <- qaptest(networks, gcor, g1 = i, g2 = j, reps = 500)
      qap_cors[i,j] <- qap_result$testval
      qap_pvals[i,j] <- qap_result$pgreq
    }
  }
}

# Display results
colnames(qap_cors) <- rownames(qap_cors) <- net_names
colnames(qap_pvals) <- rownames(qap_pvals) <- net_names

cat("\nQAP Correlation Matrix:\n")
print(round(qap_cors, 3))

cat("\nQAP P-values:\n")
print(round(qap_pvals, 3))
```

## Part 4: QAP Regression

Now let's predict advice-seeking from friendship, office proximity, and department:

```{r}
# Prepare matrices for regression
y <- advice
X <- array(dim = c(3, n, n))
X[1,,] <- friendship
X[2,,] <- office_dist  
X[3,,] <- same_dept

# Run QAP regression with OLS
qap_ols <- netlm(y, X, 
                 mode = "digraph",
                 nullhyp = "qap",
                 reps = 1000)

# Display results
cat("QAP Regression Results: Advice Network\n")
cat("=====================================\n")
summary(qap_ols)
```

::: {.callout-note}
## Understanding QAP Regression Output
- **Coefficients**: Effect of each predictor on advice-seeking
- **Std. Error**: Uncertainty in estimates
- **P-values**: Significance based on QAP permutations
- **R-squared**: Proportion of variance explained
:::

## Part 5: Visualizing Regression Results

```{r}
#| fig-width: 10
#| fig-height: 8
# Extract coefficients and create bar plot
coef_names <- c("Intercept", "Friendship", "Office Proximity", "Same Dept")
coef_vals <- qap_ols$coefficients
coef_pvals <- qap_ols$pgreqabs

# Create coefficient plot
par(mfrow = c(2, 1))

# Coefficient values
barplot(coef_vals, 
        names.arg = coef_names,
        col = ifelse(coef_pvals < 0.05, "#c41c85", "gray"),
        main = "QAP Regression Coefficients",
        ylab = "Coefficient Value",
        las = 2)
abline(h = 0, lty = 2)
legend("topright", 
       legend = c("p < 0.05", "p ≥ 0.05"),
       fill = c("#c41c85", "gray"))

# P-values
barplot(-log10(coef_pvals), 
        names.arg = coef_names,
        col = ifelse(coef_pvals < 0.05, "#50C878", "lightgray"),
        main = "Statistical Significance (-log10 p-value)",
        ylab = "-log10(p-value)",
        las = 2)
abline(h = -log10(0.05), lty = 2, col = "red")
text(4, -log10(0.05), "p = 0.05", col = "red", pos = 3)
```

## Part 6: Model Diagnostics

```{r}
# Calculate fitted values and residuals
fitted_vals <- qap_ols$fitted.values
residuals <- qap_ols$residuals

# Residual plots
par(mfrow = c(1, 2))

# Histogram of residuals
hist(residuals, 
     breaks = 30,
     main = "Distribution of Residuals",
     xlab = "Residuals",
     col = "lightblue")

# Fitted vs residuals
plot(fitted_vals, residuals,
     main = "Fitted vs Residuals",
     xlab = "Fitted Values",
     ylab = "Residuals",
     pch = 19,
     col = rgb(0, 0, 0, 0.5))
abline(h = 0, col = "red", lty = 2)
```

## Your Turn

Try these exercises:

```{r}
#| eval: false
# Exercise 1: Create and test a hypothesis
# Do people seek advice more from those in their department who are also friends?
# Hint: Create an interaction term

# Your code here


# Exercise 2: Reverse the analysis
# Predict friendship from advice-seeking, proximity, and department

# Your code here


# Exercise 3: Add a new predictor
# Create a "seniority difference" matrix and add it to the model

# Your code here
```

## Key Concepts

1. **QAP** preserves network structure during permutation tests
2. **Network autocorrelation** makes standard tests invalid
3. **Multiple regression** with QAP allows complex hypotheses
4. **Effect sizes** matter as much as significance

::: {.callout-warning}
## Common Pitfalls
- Don't use standard correlation/regression on network data
- Check for multicollinearity between predictors
- Consider directed vs undirected relationships
- Remember that QAP is computationally intensive
:::

::: {.callout-note collapse="true"}
## Solutions to Exercises

```{r}
#| eval: false
# Exercise 1 Solution: Interaction term
interaction <- friendship * same_dept
X_int <- array(dim = c(4, n, n))
X_int[1,,] <- friendship
X_int[2,,] <- office_dist
X_int[3,,] <- same_dept
X_int[4,,] <- interaction

qap_interact <- netlm(y, X_int, mode = "digraph", 
                      nullhyp = "qap", reps = 500)
summary(qap_interact)

# Exercise 2 Solution: Reverse prediction
y_friend <- friendship
X_friend <- array(dim = c(3, n, n))
X_friend[1,,] <- advice
X_friend[2,,] <- office_dist
X_friend[3,,] <- same_dept

qap_friend <- netlm(y_friend, X_friend, mode = "graph",
                    nullhyp = "qap", reps = 500)
summary(qap_friend)

# Exercise 3 Solution: Seniority difference
seniority <- c(10, 9, 8, 5, 4, 3, 3, 2, 1, 1)  # years
sen_diff <- matrix(0, n, n)
for(i in 1:n) {
  for(j in 1:n) {
    sen_diff[i,j] <- abs(seniority[i] - seniority[j])
  }
}

X_sen <- array(dim = c(4, n, n))
X_sen[1,,] <- friendship
X_sen[2,,] <- office_dist  
X_sen[3,,] <- same_dept
X_sen[4,,] <- sen_diff

qap_sen <- netlm(y, X_sen, mode = "digraph",
                 nullhyp = "qap", reps = 500)
summary(qap_sen)
```
:::