---
title: "Week 3"
subtitle: "Network centrality - case discussion"
callout-appearance: "default"
---

## Prepare

::: {.callout-note}
## ðŸ“š Prepare

ðŸ“– Review [Week 2 slideshows](../week-2/main.qmd#participate) on network centrality

ðŸ“– Revise centrality measures: degree, closeness, betweenness, and eigenvector centrality

:::

## Participate

::: {.callout-note}
## ðŸ‘¥ Participate

ðŸ™‹ **Case Discussion**: Network Centrality in Practice

This week we focus on applying centrality concepts through case study discussion. No new slideshows this week.

:::

## Practice

::: {.callout-note}
## ðŸ’» Practice

### Twitch Network Exploratory Data Analysis

Explore the Twitch streamer network dataset to understand node attributes before applying centrality measures.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r load-libraries}
# Load required libraries
library(tidyverse)
library(ggplot2)
library(scales)
library(knitr)

# Set theme for plots
theme_set(theme_minimal())
```

```{r load-data}
# Read the Twitch nodes data
twitch_nodes <- read_csv("../../../data/twitch/Twitch_nodes.csv")

# Display data structure
cat("Dataset dimensions:", nrow(twitch_nodes), "rows,", ncol(twitch_nodes), "columns\n")
```

### Data Overview

The Twitch dataset contains information about streamers including:
- **ID**: Streamer identifier
- **Days_active**: Number of days the streamer has been active
- **Mature_content**: Whether the channel contains mature content (TRUE/FALSE)
- **Views**: Total number of views
- **Affiliate**: Whether the streamer has affiliate status (TRUE/FALSE)
- **Channel_ID**: Unique channel identifier

```{r data-summary}
# Create a summary table of data types
data_summary <- data.frame(
  Variable = names(twitch_nodes),
  Type = sapply(twitch_nodes, class),
  Missing = sapply(twitch_nodes, function(x) sum(is.na(x))),
  Unique = sapply(twitch_nodes, function(x) length(unique(x))),
  stringsAsFactors = FALSE
)

kable(data_summary, caption = "Data Structure Summary")
```

### Distribution Analysis

#### Views Distribution

```{r views-analysis}
# Calculate summary statistics
views_summary <- twitch_nodes %>%
  summarise(
    Min = min(Views),
    Q1 = quantile(Views, 0.25),
    Median = median(Views),
    Mean = mean(Views),
    Q3 = quantile(Views, 0.75),
    Max = max(Views)
  )

kable(views_summary, caption = "Views Summary Statistics", format.args = list(big.mark = ","))

# Create histogram with log scale
p1 <- ggplot(twitch_nodes, aes(x = Views)) +
  geom_histogram(bins = 50, fill = "#9146FF", alpha = 0.8, color = "white") +
  scale_x_log10(labels = comma) +
  geom_vline(aes(xintercept = median(Views)), 
             color = "red", linetype = "dashed", size = 1) +
  labs(title = "Distribution of Channel Views (Log Scale)",
       subtitle = paste("Median:", format(median(twitch_nodes$Views), big.mark = ",")),
       x = "Views (log scale)",
       y = "Number of Streamers") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5))

p1
```

#### Categorical Variables

```{r categorical-analysis}
# Create summary tables
categorical_summary <- bind_rows(
  twitch_nodes %>%
    count(Mature_content) %>%
    mutate(Variable = "Mature Content",
           Category = as.character(Mature_content),
           Percentage = round(n / sum(n) * 100, 1)),
  
  twitch_nodes %>%
    count(Affiliate) %>%
    mutate(Variable = "Affiliate Status",
           Category = as.character(Affiliate),
           Percentage = round(n / sum(n) * 100, 1))
) %>%
  select(Variable, Category, n, Percentage)

kable(categorical_summary, caption = "Categorical Variables Distribution")
```

### Relationship Analysis

#### Affiliate Status vs Views

```{r affiliate-views}
p2 <- twitch_nodes %>%
  mutate(Affiliate = factor(Affiliate, 
                           levels = c(FALSE, TRUE),
                           labels = c("Non-Affiliate", "Affiliate"))) %>%
  ggplot(aes(x = Affiliate, y = Views, fill = Affiliate)) +
  geom_violin(alpha = 0.6) +
  geom_boxplot(width = 0.2, alpha = 0.8) +
  scale_y_log10(labels = comma) +
  scale_fill_manual(values = c("#9146FF", "#1DB954")) +
  labs(title = "Views Distribution by Affiliate Status",
       x = "Affiliate Status",
       y = "Views (log scale)") +
  theme(legend.position = "none",
        plot.title = element_text(hjust = 0.5, face = "bold"))

p2
```

#### Days Active vs Views

```{r scatter-analysis}
p3 <- ggplot(twitch_nodes, aes(x = Days_active, y = Views)) +
  geom_point(aes(color = Affiliate, shape = Mature_content), 
             alpha = 0.6, size = 2) +
  scale_y_log10(labels = comma) +
  scale_color_manual(values = c("#9146FF", "#1DB954"),
                     labels = c("Non-Affiliate", "Affiliate")) +
  scale_shape_manual(values = c(16, 17),
                     labels = c("Non-Mature", "Mature")) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.2) +
  labs(title = "Relationship between Days Active and Views",
       x = "Days Active",
       y = "Views (log scale)",
       color = "Affiliate Status",
       shape = "Content Type") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

p3
```

### Cross-tabulation Analysis

```{r cross-analysis}
# Create detailed cross-tabulation
cross_analysis <- twitch_nodes %>%
  group_by(Mature_content, Affiliate) %>%
  summarise(
    Count = n(),
    Avg_Views = mean(Views),
    Median_Views = median(Views),
    Avg_Days_Active = mean(Days_active),
    .groups = "drop"
  ) %>%
  mutate(
    Avg_Views = round(Avg_Views),
    Avg_Days_Active = round(Avg_Days_Active),
    Mature_content = ifelse(Mature_content, "Mature", "Non-Mature"),
    Affiliate = ifelse(Affiliate, "Affiliate", "Non-Affiliate")
  )

kable(cross_analysis, 
      caption = "Average Metrics by Content Type and Affiliate Status",
      format.args = list(big.mark = ","))
```

### Statistical Tests

```{r t-tests}
# Affiliate status comparison
affiliate_test <- t.test(log10(Views + 1) ~ Affiliate, data = twitch_nodes)
cat("T-test: Affiliate vs Non-Affiliate (log-transformed views)\n")
cat("t =", round(affiliate_test$statistic, 4), "\n")
cat("p-value =", format.pval(affiliate_test$p.value), "\n")
cat("Mean difference =", round(diff(affiliate_test$estimate), 4), "log units\n")
```

### Top Performers

```{r top-channels}
# Identify top 10 channels by views
top_channels <- twitch_nodes %>%
  arrange(desc(Views)) %>%
  slice_head(n = 10) %>%
  mutate(
    Rank = row_number(),
    Views = format(Views, big.mark = ","),
    Content_Type = ifelse(Mature_content, "Mature", "Non-Mature"),
    Status = ifelse(Affiliate, "Affiliate", "Non-Affiliate")
  ) %>%
  select(Rank, Channel_ID, Views, Days_active, Content_Type, Status)

kable(top_channels, caption = "Top 10 Twitch Channels by Views")
```

### Key Insights

```{r insights}
# Calculate key metrics
insights <- list(
  total_streamers = nrow(twitch_nodes),
  avg_views = mean(twitch_nodes$Views),
  median_views = median(twitch_nodes$Views),
  affiliate_rate = mean(twitch_nodes$Affiliate) * 100,
  mature_content_rate = mean(twitch_nodes$Mature_content) * 100,
  views_ratio = mean(twitch_nodes$Views) / median(twitch_nodes$Views)
)
```

**Summary of Key Findings:**

- **Total streamers analyzed**: `r format(insights$total_streamers, big.mark = ",")`
- **Highly skewed viewership**: Average views (`r format(round(insights$avg_views), big.mark = ",")`) are `r round(insights$views_ratio, 1)`x higher than median views (`r format(insights$median_views, big.mark = ",")`), indicating extreme concentration
- **Affiliate rate**: Only `r round(insights$affiliate_rate, 1)`% of streamers have affiliate status
- **Mature content**: `r round(insights$mature_content_rate, 1)`% of channels are marked as mature
- **Affiliate advantage**: Affiliate streamers have significantly higher viewership (p < 0.001)

This analysis provides context for understanding how node attributes might relate to network centrality measures in the Twitch streamer ecosystem.

:::

## Perform

::: {.callout-note}
## ðŸ“ Perform

âœï¸ Analyze the 'Twitch' case study according to a network centrality approach
(refer to the case studies section of SMM638's Moodle page).

ðŸ’¡ Resources on how to discuss business case studies: [The Case Method](https://www.hbs.edu/mba/academic-experience/the-case-method)

:::

## Ponder

**Network centrality applications**

Borgatti, S. P. (2005). [*Centrality and network flow*](https://doi.org/10.1016/j.socnet.2004.11.008). Social Networks, 27(1), 55-71.

**Identifying key players in networks**

Borgatti, S. P. (2006). [*Identifying sets of key players in a social network*](https://doi.org/10.1007/s10588-006-7084-x). Computational & Mathematical Organization Theory, 12(1), 21-34.

**Centrality in organizational networks**

Sparrowe, R. T., Liden, R. C., Wayne, S. J., & Kraimer, M. L. (2001). [*Social networks and the performance of individuals and groups*](https://doi.org/10.5465/amj.2001.4378755). Academy of Management Journal, 44(2), 316-325.
